---
title: "W12. Statistical models of networks"
format: 
  revealjs:
    theme: default
    slide-number: true
    chalkboard: true
execute:
  echo: true
  engine: knitr
  fig-align: center
  fig-asp: null
---

# Statistical models of networks 
- Focus on the method ERGM(exponential-family random graph models)
- Apply the model toa small school-based network(at one time point)
  - Do boys tend to have more friends than girls? 
  - Is gender homophily a strong driver of tie formation? 


## Setting up the Session
```{r message=F, warning=F}
library(ergm)
library(sna)
library(car)
```

## Setting up the Session
- Read in our example network data(high school friendship network)
```{r}

url1 <- "https://github.com/JeffreyAlanSmith/Integrated_Network_Science/raw/master/data/example_schoolmat.csv"
# adjacency matrix format
school_mat <- read.csv(file = url1, row.names = 1) 
school_mat <- as.matrix(school_mat)
school_mat[1:5, 1:5] 
```

## Setting up the Session
- Read in the attribute file
```{r}
url2 <- "https://github.com/JeffreyAlanSmith/Integrated_Network_Science/raw/master/data/example_schooldata.csv"

school_attributes <- read.csv(file = url2) 
head(school_attributes)
```


## Setting up the Session
- Recode the 0s and 1s into male and female
```{r}
school_attributes$gender_recode <- recode(school_attributes$gender, 
                                          as.factor = F, 
                                          "0 = 'male'; 1 = 'female'")
attribute_list <- do.call(list, school_attributes) 
```

## Setting up the Session
- Construct a network(with network package)

```{r}
school_net <- network(x = school_mat, directed = T, 
                      vertex.attr = attribute_list) 
school_net
```

## Descriptive Statistics
```{r fig.height=7.5, fig.width=7.5}
plot(school_net)
```

## Descriptive Statistics
- Strong gender homophily 

```{r fig.height=7.5, fig.width=7.5}
cols <- recode(school_attributes$gender, as.factor = F,
               "0 = 'blue'; 1 = 'pink'")
plot(school_net, vertex.col = cols) 
```

## Descriptive Statistics
- and weaker gender homophily

```{r fig.height=7.5, fig.width=7.5}
plot(school_net, vertex.col = school_attributes$grade) 
```



## Fitting an ERGM
- ERGM(Exponential Random Graph Model)
- Predicts the presence/absence of a tie between all i-j pairs 
- Interpret it as a network features like reciprocity and homophily

## Fitting an ERGM
- formula = the set of terms we want to include in the model. 
- control = list of options used in the algorithm
- constraints = a formula dictating if there should be any constraints on the networks generated in the MCMC sample

## Simple ERGM
- Run an ERGM with only edges as a predictor
```{r, message=F, results='hide'}
mod_rand <- ergm(formula = school_net ~ edges) 
```

## Simple ERGM
- Let’s take a look at the model results. 
```{r}
summary(mod_rand) 
```


## 1) Add options to ERGM
- Let's add `nodematch` terms for gender and grade
  - Count the number of times that an edge exists where i and j have the same attribute
```{r message=F, warning=F}
mod_homoph1 <- ergm(school_net ~ edges + nodematch("gender_recode") + 
                      nodematch("grade")) 
```

## 1) Add options to ERGM 
```{r}
summary(mod_homoph1)
```

## 2) Add options to ERGM
-  Whether boys (compared to girls) send out and/or receive more ties
  - Add `nodeifactor` terms which capture if the mean number of ties coming in 
  - Add `nodeofactor` terms which capture if the mean number of ties going out  
  - Add analogous terms for grade using `nodeicov` and `nodeocov`

```{r message=F, warning=F}
mod_homoph2 <- ergm(school_net ~ edges + nodematch("gender_recode") +
                      nodematch("grade") + 
                      nodeifactor("gender_recode") + 
                      nodeofactor("gender_recode") +
                      nodeicov("grade") + nodeocov("grade"))
```

## 2) Add options to ERGM
```{r}
summary(mod_homoph2) 
```

## 3) Add options to ERGM
- Add  a mutual term to the mode to identify reciprocity
  - `mutual` term counts the number of pairs where i->j and j->i exist. 
  - MCMC estimation : useful to set the input parameters to the estimation routine
  
```{r message=F, warning=F}
set.seed(1012)

mod_homoph_mutual1 <- ergm(school_net ~ edges + 
                             nodematch("gender_recode") + 
                             nodematch("grade") + 
                             nodeifactor("gender_recode") + 
                             nodeofactor("gender_recode") + 
                             nodeicov("grade") + nodeocov("grade") +
                             mutual, 
                           control = control.ergm(MCMC.burnin = 50000,
                                                  MCMC.samplesize = 6000)) 
```

## 4) Checking Model Convergence
- MCMC... Make sure that the algorithm converged
- Let's see how the model looks using a `mcmc.diagnostics()` function:

```{r results='hide'}
pdf("mcmc_diagnostics.pdf", width = 10, height = 8)
mcmc.diagnostics(mod_homoph_mutual1, vars.per.page = 4)
dev.off()

```

The function plots the MCMC sample of network statistics for each term included in the model. The plots show how far the statistics calculated on the sampled networks are from the true value, calculated on the observed network. The plots on the left hand show the statistics taken from each individual sample (plotted in order), while the right hand side shows the distribution of sample statistics. A 'nice' looking plot would have the sample statistics centered around 0, with some (but not extreme) variance sample to sample, suggesting that the specified model is well-behaved and can produce networks consistent with the observed data.

## 5) Checking Statistical Result 

```{r message=F}
summary(mod_homoph_mutual1)
```


## 5) Add options to ERGM again
- Push this intuition a little further 
  - If boys have more reciprocated ties than girls, and thus are involved in more mutual dyads

```{r message=F}
mod_homoph_mutual1b <- ergm(school_net ~ edges + 
                              nodematch("gender_recode") +
                              nodematch("grade") + 
                              nodeicov("grade") + nodeocov("grade") +
                              #added!
                              mutual(by = "gender_recode"),
                            control = control.ergm(MCMC.burnin = 50000, 
                                                   MCMC.samplesize = 6000))
```

## 5) Add options to ERGM again
```{r}
summary(mod_homoph_mutual1b)
```

## 6) Add options to ERGM again
- Include a control for degree differences (by gender) in the model
  - If boys are more likely to be involved in mutual dyads because they send out more ties or because expectations of reciprocity are higher among boys

```{r message=F}
mod_homoph_mutual1c <- ergm(school_net ~ edges +
                              nodematch("gender_recode") + 
                              nodematch("grade") + 
                              #added!
                              nodeofactor("gender_recode") + 
                              nodeicov("grade") + nodeocov("grade") + 
                              mutual(by = "gender_recode"),
                            control = control.ergm(MCMC.burnin = 50000, 
                                                   MCMC.samplesize = 6000))
```

## 6) Add options to ERGM again
```{r}
summary(mod_homoph_mutual1c)
```
## Goodness of Fit
- Let's see if the specified model is fitting the network well
  - Test whether the micro-level tie formation processes assumed in the model are sufficient to reproduce the features of the network as a whole 

## Goodness of Fit
- Check if our model can reproduce distance and the shared partner distribution from the true network
  - Distance captures the distribution of shortest paths between all ij pairs
```{r}
summary(school_net ~ esp(0:10)) 
```
## Goodness of Fit
-  `gof()` function will simulate a set of networks based on the estimated model
- Take the generated networks, calculate the macro features of interest (here distance and the shared partner distribution) 

```{r}
gof_mod_homoph_mutual1 <- gof(mod_homoph_mutual1, 
                              GOF = ~ distance + espartners + model,
                              control = control.gof.ergm(seed = 110)) 
```

## Goodness of Fit
- plot the goodness of fit statistics.

```{r}
par(mfrow = c(1, 3)) 
plot(gof_mod_homoph_mutual1)
```

## Adding GWESP to the Model
- Add a term to the model that will capture the tendency for nodes who are tied to have the same friends
- Capture clustering using a triangle term
  - A triangle is defined as all cases where i-j and j-k exist and then either k->i or i->k exist

```{r message=F, warning=F}
mod_homoph_mutual_gwesp1 <- ergm(school_net ~ edges + 
                                   nodematch("gender_recode") + 
                                   nodematch("grade") + 
                                   nodeifactor("gender_recode") + 
                                   nodeofactor("gender_recode") + 
                                   nodeicov("grade") + nodeocov("grade") + 
                                   mutual + gwesp(decay = 0.5, fixed = T), 
                                 control = control.ergm(MCMLE.maxit = 20,      
                                                          MCMC.samplesize = 5000, 
                                                          MCMC.burnin = 25000,    
                                                          MCMC.interval = 10000,  
                                                          MCMC.effectiveSize = 500,
                                                          seed = 123
                                                        ))
                     
```

## Adding GWESP to the Model
- Looking at the diagnostics:

```{r results='hide', warning = F}
mcmc.diagnostics(mod_homoph_mutual_gwesp1)
```

## Adding GWESP to the Model
- plotting the last network from the MCMC sample.

```{r}
plot(mod_homoph_mutual_gwesp1$newnetwork) 
```


## Done!

-   Please check KLMS for the assignment of this week(til Friday)
-   If you want to study in-depth knowledge for analyzing the two-mode analysis and hierarchical clustering for correspondence analysis, please check this week \[Option\] in the assignment
